---
output:
  pdf_document: default
  html_document: default
---
---
title: "STAT 536 HW 1 Report"
author: "Ty Hawkes and Brigg Trendler"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r, include = FALSE, message = FALSE}
# Data & Packages
library(tidyverse)

# Data
df <- vroom::vroom("Rivers.csv")
```


## Abstract

## 1. Introduction: Problem Statement and Understanding

```{r, echo = FALSE, message = FALSE}
library(maps)

# Get US state boundaries
us_states <- map_data("state")

# Create the plot
ggplot() +
  geom_polygon(data = us_states, aes(x = long, y = lat, group = group), fill = NA, color = "black") +  # Add state boundaries
  geom_point(data = df, aes(x = Lon, y = Lat, color = Metric), size = 3) +  # Add points with Metric as color
  scale_color_gradient(low = "blue", high = "red") +  # Adjust the color gradient
  labs(title = "River Flow",
       x = "Longitude",
       y = "Latitude",
       color = "Metric") +
  theme_minimal()  # Use a minimal theme for the plot
```


## 2. Methodology

### 2.1 Proposed Methods

### 2.2 Model Evaluation

## 3. Results

### 3.1 What are the biggest climate / river network / human factors that impact overall river flow?
The table below shows the beta coefficients of the factors that improved the BIC of our model. Each variable in our model had a significant impact on overall river flow.

| Term                             | Category       | Estimate   | Std. Error | Pr(>|t|)    |
|----------------------------------|----------------|------------|------------|-------------|
| (Intercept)                      |                | 103.18286  | 22.61833   | 1.61e-05 ***|
| bio15                            | Climate        | -0.59865   | 0.13978    | 4.65e-05 ***|
| CumPrec05                        | Climate        | 2.20575    | 0.63710    | 0.000826 ***|
| mPDC_SomewhatExcessive           | Land Cover     | 0.42028    | 0.16888    | 0.014682 *  |
| bio18                            | Climate        | 0.73360    | 0.37259    | 0.052074 .  |
| cls2                             | Land Cover     | 317.43807  | 65.55078   | 5.37e-06 ***|
| Lon                              |                | -0.08278   | 0.02067    | 0.000128 ***|
| cls9                             | Land Cover     | -1.32608   | 0.34508    | 0.000228 ***|
| bio14                            | Climate        | -1.25629   | 0.39629    | 0.002090 ** |
| gord                             | Network        | 0.44521    | 0.11739    | 0.000271 ***|
| mPDC_ModeratelyWell              | Land Cover     | -0.19627   | 0.09066    | 0.033074 *  |
| cls8                             | Land Cover     | 2.17050    | 1.09753    | 0.051066 .  |
| cls1                             | Land Cover     | 0.17930    | 0.09525    | 0.063035 .  |

Each category had at least one significant factor affecting river flow. Climate factors included bio15 (Precipitation Seasonality (Coefficient of Variation)), CumPrec05 (Cumulative May precipitation for the watershed upstream of grdc station), bio18 (Precipitation of Warmest Quarter), and bio14 (Precipitation of Driest Month); Land cover factors included meanPercentDC_SomewhatExcessive (mean somewhat excessive drainage class), cls2 (Evergreen_Broadleaf), cls9 (Urban), meanPercentDC_ModeratelyWell (mean moderately well drained soil), cls8 (Regularly Flooded Vegetation), and cls1 (Evergreen_Dec_Needle_Trees); and network factors included gord (global stream order from stream dem (Predicted relationship with area)).

### 3.2 How well do these factors explain overall flow?

Our model does a decent job explaining the overall flow of rivers. With these selected co-variates, our model explains nearly 82% of the in-sample variance as measured by the r-squared value.

### 3.3 How predictive of overall flow are these identified factors?

We estimate that our model will do a reasonable job predicting overall flow on our-of-sample data. Using a leave-one-out cross validation, our RMSE was .4077, a significant improvement over the base intercept model with an RMSE of .8845.

## 4. Conclusions

## Teamwork

