---
output:
  pdf_document: default
  html_document: default
---
---
title: "STAT 536 HW 1 Report"
author: "Ty Hawkes and Brigg Trendler"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r, include = FALSE, message = FALSE}
# Data & Packages
library(tidyverse)

# Data
df <- vroom::vroom("Rivers.csv")
```


## Abstract

## 1. Introduction: Problem Statement and Understanding

```{r, echo = FALSE, message = FALSE}
library(maps)

# Get US state boundaries
us_states <- map_data("state")

# Create the plot
ggplot() +
  geom_polygon(data = us_states, aes(x = long, y = lat, group = group), fill = NA, color = "black") +  # Add state boundaries
  geom_point(data = df, aes(x = Lon, y = Lat, color = Metric), size = 3) +  # Add points with Metric as color
  scale_color_gradient(low = "blue", high = "red") +  # Adjust the color gradient
  labs(title = "River Flow",
       x = "Longitude",
       y = "Latitude",
       color = "Metric") +
  theme_minimal()  # Use a minimal theme for the plot
```

Rivers are responsible for carrying water and nutrients to various parts of the earth.
Many animals, including humans, rely on rivers and the life they bring to thrive in otherwise
uninhabitable areas.
In this project, we work to analyze how various factors (human, river network, and climate)
impact the overall river flow in the Rocky Mountain area.

Through an initial exploratory data analysis, we found that we have 95 covariates and
102 observations.
This means, should we create a linear regression model with every variable, the model
would likely be severely overfit, which would lead to several potential problems,
including multicollinearity, high variance in coefficients, and interpretability issues.

The data set also contains monthly averages for precipitation, temperature, and cumulative precipitation,
as well as annual averages for those same variables.
This is potentially problematic because including both the yearly and monthly
averages would introduce multicollinearity into the model.
Furthermore, some of the variables are direct linear combinations of others.
For example, the variable bio7 gives the annual temperature range, given by
subtracting the max temperature from the minimum, both of which are stored as
bio5 and bio6.
Thus we want to avoid including these in our model simultaneously.

In this project, we hope to find a predictive model for river flow based on
the given variables while simultaneously avoiding incorporating the aforementioned
problems into our model.

## 2. Methodology

### 2.1 Proposed Methods

To combat the high dimensionality of our data set, we propose two models: a linear
regression model constructed from a LASSO regularization, and a linear regression
model created from a bi-directional selection algorithm using the AIC as a scoring metric.

LASSO regularization helps reduce dimensionality selecting the most relevant predictors.
It improves interpretability by producing a sparse model and reduces overfitting.
However, it can introduce bias into coefficient estimates, and can struggle 
when trying to handle highly correlated variables.
In the context of the problem of this case study, a model built by LASSO would solve
the dimensionality problem, however the colinearity between variables might pose a problem
for this method.

A forward and backward selection algorithm balances model fit and complexity.
It also automates variable selection and has the advantage of navigating correlated
predictors better than a pure forward or pure backward selection algorithm.
Unfortunately, it is a greedy algorithm, meaning it may not find the best
model.
It also does not handle multicollinearity well and can be computationally expensive.
A forward and backward selection model in the context of this problem would also
solve the dimensionality issue.

In both cases, we will need to check the classic LINE assumptions when using
linear regression: linearity, independence, normally distributed errors, and equal variance.

### 2.2 Model Evaluation

Both models seemed to take a conservative approach to interpretability by
selecting the key components and favoring model simplicity.
We decided to test both models using leave-one-out cross validation and found
that the forward-and-backward selection algorithm model appeared to perform better,
both in sample and out of sample.

To verify our model is operating under the correct assumptions, we plotted
the residuals vs fitted values graph and found that the residuals are randomly
scattered around zero, indicating that has mostly captured the true relationship
between the predictors and the river flow metric. This also indicates that
the assumptions of linearity and constant variance are likely met.

We generated a normal Q-Q plot and found that the points fall close to the diagonal
line, which suggests that the data likely follows from a normal distribution,
again validating one of our underlying assumptions when working with linear
regression models.

The scale-location plot reveals a downward trend of our standardized residuals.
This indicates that there is a bit of heteroscedasticity in the data.
However, the part that appears to be heteroscedastic also has the fewest
data points.
The bulk of the residuals appear to be clustered around 0, which may indicate
that the data is homoscedastic enough for our purposes.

We also checked the residuals vs leverage and find that there are no
blatant problems.
There are no distinct patterns, meaning the model fits well even in the presence
of the outliers in the data set.

Finally, we made sure our final model did not contain colinear variables
(such as the yearly averages and bio7) as the information contained in those variables
were also found in other variables or was deemed not as important by the feature
selection algorithm.


| Model | $R^2$ | RMSE |
|-------|-------|------|
| LASSO | 0.8065 | 0.484 |
| Forward and Backward Selection | 0.8297 | 0.4077 |

## 3. Results

### 3.1 What are the biggest climate / river network / human factors that impact overall river flow?
The table below shows the beta coefficients of the factors that improved the AIC of our model. Each variable in our model had a significant impact on overall river flow.

| Term                             | Category       | Estimate   | Std. Error | Pr(>|t|)    |
|----------------------------------|----------------|------------|------------|-------------|
| (Intercept)                      |                | 103.18286  | 22.61833   | 1.61e-05 ***|
| bio15                            | Climate        | -0.59865   | 0.13978    | 4.65e-05 ***|
| CumPrec05                        | Climate        | 2.20575    | 0.63710    | 0.000826 ***|
| mPDC_SomewhatExcessive           | Land Cover     | 0.42028    | 0.16888    | 0.014682 *  |
| bio18                            | Climate        | 0.73360    | 0.37259    | 0.052074 .  |
| cls2                             | Land Cover     | 317.43807  | 65.55078   | 5.37e-06 ***|
| Lon                              |                | -0.08278   | 0.02067    | 0.000128 ***|
| cls9                             | Land Cover     | -1.32608   | 0.34508    | 0.000228 ***|
| bio14                            | Climate        | -1.25629   | 0.39629    | 0.002090 ** |
| gord                             | Network        | 0.44521    | 0.11739    | 0.000271 ***|
| mPDC_ModeratelyWell              | Land Cover     | -0.19627   | 0.09066    | 0.033074 *  |
| cls8                             | Land Cover     | 2.17050    | 1.09753    | 0.051066 .  |
| cls1                             | Land Cover     | 0.17930    | 0.09525    | 0.063035 .  |

Each category had at least one significant factor affecting river flow. Climate factors included bio15 (Precipitation Seasonality (Coefficient of Variation)), CumPrec05 (Cumulative May precipitation for the watershed upstream of grdc station), bio18 (Precipitation of Warmest Quarter), and bio14 (Precipitation of Driest Month); Land cover factors included meanPercentDC_SomewhatExcessive (mean somewhat excessive drainage class), cls2 (Evergreen_Broadleaf), cls9 (Urban), meanPercentDC_ModeratelyWell (mean moderately well drained soil), cls8 (Regularly Flooded Vegetation), and cls1 (Evergreen_Dec_Needle_Trees); and network factors included gord (global stream order from stream dem (Predicted relationship with area)).

### 3.2 How well do these factors explain overall flow?

Our model does a decent job explaining the overall flow of rivers. With these selected co-variates, our model explains nearly 82% of the in-sample variance as measured by the r-squared value.

### 3.3 How predictive of overall flow are these identified factors?

We estimate that our model will do a reasonable job predicting overall flow on our-of-sample data. Using a leave-one-out cross validation, our RMSE was .4077, a significant improvement over the base intercept model with an RMSE of .8845.

## 4. Conclusions

## Teamwork

